Comparison of particle trajectories and collision operators for collisional transport in nonaxisymmetric plasmas

### 2014-02-18 doc/sfincsPaper/sfincsPaper.tex
Finishing discussion of momentum balance in revised manuscript

As shown in Appendix C of Ref. \cite{usPedestal}, we may choose the gauge
for the electromagnetic potentials such that $-c^{-1}\vect{b}\cdot\partial\vect{A}/\partial t = \left<E_{||} B\right> B/\left<B^2\right>$
on the right-hand side of (\ref{eq:kinetic1}).
Here, angle brackets denote a flux surface average:
\begin{equation}
\label{eq:FSA}
\left\langle \ldots \right\rangle
=\frac{1}{{V}'}
\int_0^{2\pi }d\theta
\int_0^{2\pi }d\zeta
{\frac{(\ldots)}{\vect{B}\cdot \nabla \zeta }}
\end{equation}
where
${V}'=\int_0^{2\pi } d\theta \int_0^{2\pi } d\zeta / \vect{B}\cdot \nabla \zeta $,
$\theta$ and $\zeta$ are poloidal and toroidal magnetic angles satisfying
\begin{equation}
\vect{B} = \nabla \psi \times \nabla \theta + \iota \nabla\zeta\times\nabla\psi,
\end{equation}
$\iota = 1/q$ is the rotational transform, and $q$ is the safety factor.
Thus, (\ref{eq:kinetic1}) becomes
\begin{equation}
\left(\vpar \vect{b} + \vE + \vma\right) \cdot\left(\nabla f_{a1}\right)_{W_a,\mu}
-C_{a}
=-\left(\vma + \vE\right)\cdot\nabla\psi \left(\frac{\partial f_{a0}}{\partial \psi}\right)_{W_a}
+\frac{Z_a e}{T_a}\vpar\frac{B\left<E_{||} B\right>}{\left< B^2\right>} f_{a0}.
\label{eq:kinetic2}
\end{equation}
Even if the radial electric field is considered an input, this form of the kinetic equation remains
nonlinear in the unknowns since the $\nabla f_{a1}$ term depends on the variation of $\Phi$
on a flux surface, and this variation is an unknown like $f_{a1}$.

@@

\changed{
Nonetheless, it is interesting to compare the $\int d^3v\, m_a v_{||} (\ldots)$ moment equations
for each trajectory model to the full parallel momentum fluid equation. This later equation is
\begin{equation}
???
\end{equation}
First, consider the case of no radial electric field.
The stress tensor is given to the accuracy needed by $\tens{\Pi}_a \approx \tens{\Pi}_{a||}$
where $\tens{\Pi}_{a||} = p_{a \bot} \tens{I} + (p_{a||} - p_{a\bot}) \vect{b}\vect{b}$,
$p_{a||} = m_a \int d^3v\, f_{a1} v_{||}^2$, and $p_{a\bot} = m_a \int d^3v\, f_{a1} v_{\bot}^2/2$.
Then noting the $\int d^3v\, m_a v_{||} (\ldots)$ moment of the streaming and mirror terms is
\begin{eqnarray}
\int d^3v\, m_a v_{||} \left[ v_{||} \vect{b}\cdot\nabla f_{a1} - \frac{1-\xi^2}{2B}v(\vect{b}\cdot\nabla B) \frac{\partial f_{a1}}{\partial \xi} \right]
& = &
\vect{b}\cdot\nabla p_{a||1} + \frac{ p_{a\bot 1} - p_{a || 1}}{B} \vect{b}\cdot\nabla B \nonumber \\
& = &
 -\vect{b} \cdot \left(\nabla \cdot \tens{\Pi}_{a||} \right),
\end{eqnarray}
one can see the drift-kinetic equation (\ref{}) is consistent with the full fluid parallel momentum equation
at least when $E_r=0$.
}

\changed{
Now consider how the situation changes when a radial electric field is introduced. We first
compute the change to the fluid parallel momentum equation caused by a new contribution
to the viscosity.
Examining the $\left( \vect{E} + c^{-1} \vect{v}\times\vect{B}\right) \cdot \nabla_{\vect{v}} f_a$ terms in 
the full Fokker-Planck equation and integrating in gyrophase, one sees the gyrophase-dependent part of the 
distribution function $\tilde{f}_a$ will include the following terms proportional to the electric field:
\begin{equation}
\tilde{f}_{aE} = -\frac{c}{B} \vect{v}\cdot\vect{b}\times\nabla\Phi \left[ \frac{1}{v} \frac{\partial \bar{f}_a}{\partial v}
-\frac{\xi}{v^2} \frac{\partial \bar{f}_a}{\partial \xi} \right]
\label{eq:fTilde}
\end{equation}
as reflected (using different indepedent variables) in eq (17) of Ref. \cite{Hazeltine} and eq (6) of Ref. \cite{SimakovCattoSecondOrderDKE}.
Here, $\bar{f}_a$ is the gyrophase-independent part of the distribution function.
The associated contribution $\tens{\Pi}_{aE} = m_a \int d^3v\, \tilde{f}_{aE}\vect{v}\vect{v}$
to the pressure tensor is calculated in equations (27)-(36) of Ref. \cite{SimakovCattoSecondOrderDKE},
with the result
\begin{equation}
\tens{\Pi}_{aE} = m_a n_a V_{a||} \left( \vect{b}\vect{v}_E + \vect{v}_E \vect{b} \right)
\label{eq:PiE}
\end{equation}
where $n_a V_{a||} = \int d^3v\, \bar{f}_a v_{||}$. Note that this contribution to the stress tensor is a part of the gyroviscosity and is off-diagonal.
Using $\Phi \approx \Phi_0(\psi)$ and $\vect{B} \cdot \left\{ \nabla \cdot \left[ \vect{B}\vect{B}\times\nabla\psi + (\vect{B}\times\nabla\psi) \vect{B} \right] \right\}
=2B^2\nabla\psi\cdot\nabla\times\vect{B}=0$,
we then find the contribution to the parallel momentum equation from (\ref{eq:fTilde})-(\ref{eq:PiE}) is
\begin{equation}
\vect{b}\cdot(\nabla\cdot\tens{\Pi}_{aE}) = 
c m_a B (d\Phi_0/d\psi)
\vect{B}\times\nabla\psi\cdot\nabla\left(n_a V_{a||}/B^3\right).
\label{eq:PiEParallel}
\end{equation}
}

\changed{
Finally, let us consider the $\int d^3v\, m_a v_{||} (\ldots)$ moment of the radial electric field
terms in the drift-kinetic equation for various trajectory models, to see if the results agree with (\ref{eq:PiEParallel}).
For the full trajectories, the moment of the $d\Phi_0/d\psi$ terms in (\ref{eq:kinetic4})-(\ref{eq:fullTrajectories}) is
\begin{eqnarray}
\int d^3v\, m_a v_{||} \left[ 
\frac{c}{B^2} \frac{d\Phi_0}{d\psi} \vect{B}\times\nabla\psi\cdot\nabla f_{a1}
- (\vma \cdot\nabla\psi) \frac{Z_a e}{2 T_a x_a} \frac{d\Phi_0}{d\psi} \frac{\partial f_{a1}}{\partial x_a}
\right.
\nonumber \\
\left.
+ \xi(1-\xi^2) \frac{c}{2B^3} \frac{d\Phi_0}{d\psi} (\vect{B}\times\nabla\psi\cdot\nabla B)\frac{\partial f_{a}}{\partial \xi}
\right]
\nonumber \\
= 
c m_a B (d\Phi_0/d\psi)
\vect{B}\times\nabla\psi\cdot\nabla\left(n_a V_{a||}/B^3\right),
\label{eq:fullViscosity}
\end{eqnarray}
obtained by integrating by parts in $x_a$ and $\xi$.
Thus, the full-trajectory model agrees with the full fluid parallel momentum equation: (\ref{eq:fullViscosity}) = (\ref{eq:PiEParallel}).
However, this agreement is
not shared by the DKES model: the moment of the $d\Phi_0/d\psi$ term in (\ref{eq:DKESTrajectories}) is
\begin{eqnarray}
\int d^3v\, m_a v_{||} \left[ 
\frac{c}{\left< B^2 \right> } \frac{d\Phi_0}{d\psi} \vect{B}\times\nabla\psi\cdot\nabla f_{a1}
\right]
= 
\frac{c m_a}{ \left< B^2 \right>}
\frac{d\Phi_0}{d\psi}
\vect{B}\times\nabla\psi\cdot\nabla (n_a V_{a||}),
\label{eq:DKESViscosity}
\end{eqnarray}
which does not equal (\ref{eq:PiEParallel}). The corresponding result for the partial trajectories,
obtained by replacing $\left< B^2 \right> \to B^2 $ in (\ref{eq:DKESViscosity}), also does not match
(\ref{eq:PiEParallel}). Thus, the DKES and partial trajectory models do not correctly account for the parallel
viscous force as the full trajectory model does.
}

\changed{
We close this section by noting another an important difference between the trajectory models
related to the parallel momentum equations.
Consider the case of \emph{quasisymmetry}, which is the condition \cite{PerAndreiPRL} that
$\vect{B}\times\nabla\psi\cdot\nabla B = A(\psi) \vect{B}\cdot\nabla B$ for some flux function
$A(\psi)$.  It was known previously \cite{PerAndreiPRL} that when the $E_r$ terms are not included
in the trajectories (but retained in the $\partial F_a/\partial \psi$ drive term in (\ref{eq:kinetic4})), the radial neoclassical
current vanishes for all values of $d \Phi_0/d\psi$
if and only if the flux surface is quasisymmetric. This property of quasisymmetric flux surfaces
is known as intrinsic ambipolarity.
Here, we show that intrinsic ambipolarity persists in quasisymmetric geometry
when the $E_r$ terms are retained in the full trajectory
drift-kinetic equation, but not for the DKES or partial trajectory kinetic equations.
This result follows from the $-\sum_a Z_a \left< \int d^3v\, A v_{||} / \Omega_a (\ldots)\right>$
moment of the kinetic equations, i.e. a spatially weighted average of the parallel momentum
moment. For the full trajectories, the $d\Phi_0/d\psi$ terms
may be integrated by parts, canceling each other to leave
\begin{equation}
\sum_a Z_a \left< \int d^3v\, f_{a1} \vma\cdot\nabla\psi\right>=0,
\label{eq:noCurrent}
\end{equation}
meaning the radial current vanishes. However, for the DKES and partial trajectory models, the $d\Phi_0/d\psi$ terms
do not cancel, leaving an additional term in (\ref{eq:noCurrent}) proportional to $d\Phi_0/d\psi$,
meaning the radial current is generally nonzero.  Notice that when the full trajectory model is applied in quasisymmetry,
intrinsic ambipolarity means (\ref{eq:unphysicalEnergyConservation2}) is satisfied even when $d\Phi_0/d\psi \ne 0$,
so the net heat source vanishes for any radial electric field.
}

@@

Note in Figs.~(\ref{fig:collisionComparison_LHD})-(\ref{fig:collisionComparison_W7X}) that the scaling of the $L_{11}$ and $L_{12}$ coefficients at high collisionality depends crucially on whether momentum is conserved in the collision operator. In the momentum-conserving calculations, these transport coefficients are inversely proportional to $\nu$, whereas they are proportional to $\nu$ if the collisions are approximated by pure pitch-angle scattering. To understand why, it is useful to recall that the Pfirsch-Schl{\"u}ter
particle flux consists of two terms: one related to the parallel
friction force and one related to parallel viscosity \cite{AndreiPer2009, Braun}. This is most easily seen by taking the scalar product of the lowest-order plasma current, which satisfies $\vect{J} \times \vect{B} = c p'(\psi) \nabla \psi$, with the momentum equation,
\begin{equation}
m_a n_a \vect{V}_a \cdot \nabla \vect{V}_a = n_a e_a \left( - \nabla \phi + c^{-1}\vect{V}_a \times \vect{B} \right)
- \nabla p_a - \nabla \cdot \vect{\pi}_a + \vect{F}_a, 
\end{equation}
neglecting the left-hand side. Since $\nabla \cdot \vect{J} = 0$ and $n_a$ is a flux function in lowest order, we obtain
	\begin{equation}
	 \left< n_a \vect{V}_a \cdot \nabla \psi \right> 
	= \frac{1}{e_a p'(\psi)} \left< \vect{J} \cdot \left( \vect{F}_a - \nabla \cdot \vect{\pi}_a \right) \right>,
	\end{equation}
where the term corresponding to the perpendicular component of the friction force $\vect{F}_a$ represents the classical particle flux and the other terms the neoclassical flux,
		\begin{equation}
	 \left< n_a \vect{V}_a \cdot \nabla \psi \right>_{nc} = \frac{1}{e_ap'} \left< J_\| F_{a\|} + \vect{\pi}:\nabla \vect{J} \right>
	 \label{PS flux}
	 \end{equation}
where the viscosity tensor is $\vect{\pi}_a =(p_{\|} - p_{\perp})(\vect{b b} - {\bf I}/3)$.
The first term in (\ref{PS flux}) is proportional to $\nu$ and
therefore dominates at high collisionality, but vanishes when there is only a 
single ion species because of momentum conservation in like-particle
collisions. All that remains is therefore the particle flux caused
by parallel viscosity, which is inversely proportional to $\nu$ at high collisionality \cite{PerBook}. In
the pure pitch-angle-scattering model however, parallel momentum
conservation is violated, leading to spurious friction-driven
transport proportional to $\nu$. This is why the green curves have a
slope of +1 for large $\nu$ in the logarithmic plots of $L_{11}$ and
$L_{12}$ in Figures
(\ref{fig:collisionComparison_LHD})-(\ref{fig:collisionComparison_W7X}),
while the blue and red curves have the slope -1.

A similar difference between the momentum-conserving and
pitch-angle-scattering operators is evident in the parallel
conductivity coefficient $L_{33}$. The flow that arises in response to
a parallel electric field is determined by the parallel momentum equation
\begin{equation} 
n_a Z_a e E_{||} = \vect{b}\cdot \nabla p_a + \vect{b}\cdot(\nabla \cdot \vect{\pi}_a )+ F_{a||}, 
\end{equation} 
where the parallel friction force $F_{\|a}$ again vanishes when only a
single ion species is considered. Hence 
\begin{equation} 
n_a Z_a e \left< B  E_{||} \right> = \left< \vect{B}\cdot(\nabla \cdot \vect{\pi}_a) \right> 
= \left< (p_\perp - p_\| ) \nabla_\| B \right>.
\label{ParallelMomentum}
\end{equation}
In
the absence of radial gradients, the pressure anisotropy in the \PS
regime is proportional to the parallel flow velocity and inversely
proportional to the collision frequency \cite{AndreiPer2009}. The
flow $\left< V_\| B \right>$ is therefore proportional to $\nu$ in
the \PS regime unless momentum conservation is violated. In the
latter case, the spurious friction force causes $\left< V_\| B
\right>$ to be inversely proportional to $\nu$, as can be seen in
Figs.~(\ref{fig:collisionComparison_LHD})-(\ref{fig:collisionComparison_W7X}).

>>>>

As shown in Appendix C of Ref. \cite{usPedestal}, we may choose the gauge
for the electromagnetic potentials such that 
\begin{equation}
-c^{-1}\vect{b}\cdot\partial\vect{A}/\partial t = \left<E_{||} B\right> B/\left<B^2\right>
\label{eq:gauge}
\end{equation}
on the right-hand side of (\ref{eq:kinetic1}).
Here, angle brackets denote a flux surface average:
\begin{equation}
\label{eq:FSA}
\left\langle \ldots \right\rangle
=\frac{1}{{V}'}
\int_0^{2\pi }d\theta
\int_0^{2\pi }d\zeta
{\frac{(\ldots)}{\vect{B}\cdot \nabla \zeta }}
\end{equation}
where
${V}'=\int_0^{2\pi } d\theta \int_0^{2\pi } d\zeta / \vect{B}\cdot \nabla \zeta $,
$\theta$ and $\zeta$ are poloidal and toroidal magnetic angles satisfying
\begin{equation}
\vect{B} = \nabla \psi \times \nabla \theta + \iota \nabla\zeta\times\nabla\psi,
\end{equation}
$\iota = 1/q$ is the rotational transform, and $q$ is the safety factor.
Thus, (\ref{eq:kinetic1}) becomes
\begin{equation}
\left(\vpar \vect{b} + \vE + \vma\right) \cdot\left(\nabla f_{a1}\right)_{W_a,\mu}
-C_{a}
=-\left(\vma + \vE\right)\cdot\nabla\psi \left(\frac{\partial f_{a0}}{\partial \psi}\right)_{W_a}
+\frac{Z_a e}{T_a}\vpar\frac{B\left<E_{||} B\right>}{\left< B^2\right>} f_{a0}.
\label{eq:kinetic2}
\end{equation}
Even if the radial electric field is considered an input, this form of the kinetic equation remains
nonlinear in the unknowns since the $\nabla f_{a1}$ term depends on the variation of $\Phi$
on a flux surface, and this variation is an unknown like $f_{a1}$.

@@

\changed{
Nonetheless, it is interesting to compare the $\int d^3v\, m_a v_{||} (\ldots)$ moment equations
for each drift-kinetic trajectory model to the full parallel momentum fluid equation. This later equation,
the moment of the full Fokker-Planck equation, is
\begin{equation}
0 = -\vect{b}\cdot\left( \nabla\cdot\tens{\Pi}_a\right)
+ Z_a e n_a E_{||} + F_{a||}
\label{eq:fluidParallelMomentum}
\end{equation}
where $\tens{\Pi_a} = m_a \int d^3v\, f_a \vect{v}\vect{v}$ is the stress tensor
and $F_{a||}$ is the parallel component of friction.
First, consider the case of no radial electric field.
Recalling
$f_a = F_a(\psi)\left[ 1 - Z_a e \Phi_1/T_a\right] + f_{a1}$, 
the stress tensor is given to the accuracy needed by 
$\tens{\Pi}_a \approx p_a(\psi)\left[1 - Z_a e \Phi_1/T_a\right] \tens{I} + \tens{\Pi}_{a1}$
where $\tens{\Pi}_{a1} = p_{a 1 \bot} \tens{I} + (p_{a1||} - p_{a1\bot}) \vect{b}\vect{b}$,
$p_{a1||} = m_a \int d^3v\, f_{a1} v_{||}^2$, and $p_{a1\bot} = m_a \int d^3v\, f_{a1} v_{\bot}^2/2$.
Notice the $\int d^3v\, m_a v_{||} (\ldots)$ moment of the streaming and mirror terms in (\ref{eq:kinetic4})-(\ref{eq:fullTrajectories}) is
\begin{eqnarray}
\int d^3v\, m_a v_{||} \left[ v_{||} \vect{b}\cdot\nabla f_{a1} - \frac{1-\xi^2}{2B}v(\vect{b}\cdot\nabla B) \frac{\partial f_{a1}}{\partial \xi} \right]
& = &
\vect{b}\cdot\nabla p_{a1||} + \frac{ p_{a1\bot} - p_{a1 ||}}{B} \vect{b}\cdot\nabla B \nonumber \\
& = &
\vect{b} \cdot \left(\nabla \cdot \tens{\Pi}_{a1} \right).
\end{eqnarray}
Using this result with (\ref{eq:gauge}), the $m_a v_{||}$ moment of 
the drift-kinetic equation (\ref{eq:kinetic4})-(\ref{eq:fullTrajectories}) matches
the full fluid parallel momentum equation (\ref{eq:fluidParallelMomentum})
at least when $E_r=0$.
}

\changed{
Now consider how the situation changes when a radial electric field is introduced. We first
compute the change to the fluid parallel momentum equation caused by a new contribution
to the viscosity.
Examining the $\left( \vect{E} + c^{-1} \vect{v}\times\vect{B}\right) \cdot \nabla_{\vect{v}} f_a$ terms in 
the full Fokker-Planck equation and integrating in gyrophase, one sees the gyrophase-dependent part of the 
distribution function $\tilde{f}_a$ will include the following terms proportional to the electric field:
\begin{equation}
\tilde{f}_{aE} = \frac{c}{B} \vect{v}\cdot\vect{b}\times\vect{E} \left[ \frac{1}{v} \frac{\partial \bar{f}_a}{\partial v}
-\frac{\xi}{v^2} \frac{\partial \bar{f}_a}{\partial \xi} \right]
\label{eq:fTilde}
\end{equation}
as reflected (using different indepedent variables) in eq (17) of Ref. \cite{Hazeltine} and eq (6) of Ref. \cite{SimakovCattoSecondOrderDKE}.
Here, $\bar{f}_a$ is the gyrophase-independent part of the distribution function.
The associated contribution $\tens{\Pi}_{aE} = m_a \int d^3v\, \tilde{f}_{aE}\vect{v}\vect{v}$
to the pressure tensor is calculated in equations (27)-(36) of Ref. \cite{SimakovCattoSecondOrderDKE},
with the result
\begin{equation}
\tens{\Pi}_{aE} = m_a n_a V_{a||} \left( \vect{b}\vect{v}_E + \vect{v}_E \vect{b} \right)
\label{eq:PiE}
\end{equation}
where $n_a V_{a||} = \int d^3v\, f_a v_{||}$. Note that this contribution to the stress tensor is a part of the gyroviscosity and is off-diagonal.
Using $\vect{E} \approx -\nabla \Phi_0(\psi)$ and $\vect{B} \cdot \left\{ \nabla \cdot \left[ \vect{B}\vect{B}\times\nabla\psi + (\vect{B}\times\nabla\psi) \vect{B} \right] \right\}
=2B^2\nabla\psi\cdot\nabla\times\vect{B}=0$,
we then find the contribution to the parallel momentum equation from (\ref{eq:fTilde})-(\ref{eq:PiE}) is
\begin{equation}
\vect{b}\cdot(\nabla\cdot\tens{\Pi}_{aE}) = 
c m_a B (d\Phi_0/d\psi)
\vect{B}\times\nabla\psi\cdot\nabla\left(n_a V_{a||}/B^3\right).
\label{eq:PiEParallel}
\end{equation}
}

\changed{
For comparison, let us consider the $\int d^3v\, m_a v_{||} (\ldots)$ moment of the radial electric field
terms in the drift-kinetic equation for various trajectory models, to see if the results agree with (\ref{eq:PiEParallel}).
For the full trajectories, the moment of the $d\Phi_0/d\psi$ terms in (\ref{eq:kinetic4})-(\ref{eq:fullTrajectories}) is
\begin{eqnarray}
\int d^3v\, m_a v_{||} \left[ 
\frac{c}{B^2} \frac{d\Phi_0}{d\psi} \vect{B}\times\nabla\psi\cdot\nabla f_{a1}
- (\vma \cdot\nabla\psi) \frac{Z_a e}{2 T_a x_a} \frac{d\Phi_0}{d\psi} \frac{\partial f_{a1}}{\partial x_a}
\right.
\nonumber \\
\left.
+ \xi(1-\xi^2) \frac{c}{2B^3} \frac{d\Phi_0}{d\psi} (\vect{B}\times\nabla\psi\cdot\nabla B)\frac{\partial f_{a}}{\partial \xi}
\right]
\nonumber \\
= 
c m_a B (d\Phi_0/d\psi)
\vect{B}\times\nabla\psi\cdot\nabla\left(n_a V_{a||}/B^3\right),
\label{eq:fullViscosity}
\end{eqnarray}
obtained by integrating by parts in $x_a$ and $\xi$.
Thus, the full-trajectory model agrees with the full fluid parallel momentum equation: (\ref{eq:fullViscosity}) = (\ref{eq:PiEParallel}).
However, this agreement is
not shared by the DKES model: the moment of the $d\Phi_0/d\psi$ term in (\ref{eq:DKESTrajectories}) is
\begin{eqnarray}
\int d^3v\, m_a v_{||} \left[ 
\frac{c}{\left< B^2 \right> } \frac{d\Phi_0}{d\psi} \vect{B}\times\nabla\psi\cdot\nabla f_{a1}
\right]
= 
\frac{c m_a}{ \left< B^2 \right>}
\frac{d\Phi_0}{d\psi}
\vect{B}\times\nabla\psi\cdot\nabla (n_a V_{a||}),
\label{eq:DKESViscosity}
\end{eqnarray}
which does not equal (\ref{eq:PiEParallel}). The corresponding result for the partial trajectories,
obtained by replacing $\left< B^2 \right> \to B^2 $ in (\ref{eq:DKESViscosity}), also does not match
(\ref{eq:PiEParallel}). Thus, the DKES and partial trajectory models do not correctly account for the parallel
viscous force as the full trajectory model does.
}

\changed{
We close this section by noting another an important difference between the trajectory models
related to the parallel momentum equations.
Consider the case of \emph{quasisymmetry}, which is the condition \cite{PerAndreiPRL} that
$\vect{B}\times\nabla\psi\cdot\nabla B = A(\psi) \vect{B}\cdot\nabla B$ for some flux function
$A(\psi)$.  It was known previously \cite{PerAndreiPRL} that when the $E_r$ terms are not included
in the trajectories (but retained in the $\partial F_a/\partial \psi$ drive term in (\ref{eq:kinetic4})), the radial neoclassical
current vanishes for all values of $d \Phi_0/d\psi$
if and only if the flux surface is quasisymmetric. This property of quasisymmetric flux surfaces
is known as intrinsic ambipolarity.
Here, we show that intrinsic ambipolarity persists in quasisymmetric geometry
when the $E_r$ terms are retained in the full trajectory
drift-kinetic equation, but not for the DKES or partial trajectory kinetic equations.
This result follows from the $-\sum_a Z_a \left< \int d^3v\, A v_{||} / \Omega_a (\ldots)\right>$
moment of the kinetic equations, i.e. a spatially weighted average of the parallel momentum
moment. For the full trajectories, 
(\ref{eq:fullViscosity}) vanishes in this spatial average, leaving
\begin{equation}
\sum_a Z_a \left< \int d^3v\, f_{a1} \vma\cdot\nabla\psi\right>=0,
\label{eq:noCurrent}
\end{equation}
meaning there is no radial current. However, for the DKES and partial trajectory models, the
spatial average does not annihilate the $d\Phi_0/d\psi$ terms,
leaving an additional term in (\ref{eq:noCurrent}) proportional to $d\Phi_0/d\psi$,
and therefore the radial current is generally nonzero.  
Consequently, the full trajectory model is the only one of the models that preserves
intrinsic ambipolarity in quasisymmetry for $E_r \ne 0$.
Notice that when the full trajectory model is applied in quasisymmetry,
intrinsic ambipolarity means (\ref{eq:unphysicalEnergyConservation2}) is satisfied even when $d\Phi_0/d\psi \ne 0$,
so the net heat source vanishes for any radial electric field.
}

@@

Note in Figs.~(\ref{fig:collisionComparison_LHD})-(\ref{fig:collisionComparison_W7X}) that the scaling of the $L_{11}$ and $L_{12}$ coefficients at high collisionality depends crucially on whether momentum is conserved in the collision operator. In the momentum-conserving calculations, these transport coefficients are inversely proportional to $\nu$, whereas they are proportional to $\nu$ if the collisions are approximated by pure pitch-angle scattering. To understand why, it is useful to recall that the Pfirsch-Schl{\"u}ter
particle flux consists of two terms: one related to the parallel
friction force and one related to parallel viscosity \cite{AndreiPer2009, Braun}. This is most easily seen by taking the scalar product of the lowest-order plasma current, which satisfies $\vect{J} \times \vect{B} = c p'(\psi) \nabla \psi$, with the momentum equation,
\begin{equation}
m_a n_a \vect{V}_a \cdot \nabla \vect{V}_a = n_a e_a \left( - \nabla \phi + c^{-1}\vect{V}_a \times \vect{B} \right)
- \nabla p_a - \nabla \cdot \vect{\pi}_a + \vect{F}_a, 
\label{eq:vectorMomentum}
\end{equation}
neglecting the left-hand side. Since $\nabla \cdot \vect{J} = 0$ and $n_a$ is a flux function in lowest order, we obtain
\begin{equation}
  \left< n_a \vect{V}_a \cdot \nabla \psi \right> 
  = \frac{1}{e_a p'(\psi)} \left< \vect{J} \cdot \left( \vect{F}_a - \nabla \cdot \vect{\pi}_a \right) \right>,
\end{equation}
where the term corresponding to the perpendicular component of the friction force $\vect{F}_a$ represents the classical particle flux and the other terms the neoclassical flux,
\begin{equation}
  \left< n_a \vect{V}_a \cdot \nabla \psi \right>_{nc} = \frac{1}{e_ap'} \left< J_\| F_{a\|} + \vect{\pi}:\nabla \vect{J} \right>
  \label{PS flux}
\end{equation}
where the viscosity tensor is $\vect{\pi}_a =(p_{\|} - p_{\perp})(\vect{b b} - {\bf I}/3)$.
The first term in (\ref{PS flux}) is proportional to $\nu$ and
therefore dominates at high collisionality, but vanishes when there is only a 
single ion species because of momentum conservation in like-particle
collisions. All that remains is therefore the particle flux caused
by parallel viscosity, which is inversely proportional to $\nu$ at high collisionality \cite{PerBook}. In
the pure pitch-angle-scattering model however, parallel momentum
conservation is violated, leading to spurious friction-driven
transport proportional to $\nu$. This is why the green curves have a
slope of +1 for large $\nu$ in the logarithmic plots of $L_{11}$ and
$L_{12}$ in Figures
(\ref{fig:collisionComparison_LHD})-(\ref{fig:collisionComparison_W7X}),
while the blue and red curves have the slope -1.

A similar difference between the momentum-conserving and
pitch-angle-scattering operators is evident in the parallel
conductivity coefficient $L_{33}$. The flow that arises in response to
a parallel electric field is determined by the parallel momentum equation $\vect{b}\cdot$(\ref{eq:vectorMomentum}),
%\begin{equation} 
%n_a Z_a e E_{||} = \vect{b}\cdot \nabla p_a + \vect{b}\cdot(\nabla \cdot \vect{\pi}_a )+ F_{a||}, 
%\end{equation} 
where the parallel friction force $F_{\|a}$ again vanishes when only a
single ion species is considered. Hence 
\begin{equation} 
n_a Z_a e \left< B  E_{||} \right> = \left< \vect{B}\cdot(\nabla \cdot \vect{\pi}_a) \right> 
= \left< (p_\perp - p_\| ) \nabla_\| B \right>.
\label{ParallelMomentum}
\end{equation}
In
the absence of radial gradients, the pressure anisotropy in the \PS
regime is proportional to the parallel flow velocity and inversely
proportional to the collision frequency \cite{AndreiPer2009}. The
flow $\left< V_\| B \right>$ is therefore proportional to $\nu$ in
the \PS regime unless momentum conservation is violated. In the
latter case, the spurious friction force causes $\left< V_\| B
\right>$ to be inversely proportional to $\nu$, as can be seen in
Figs.~(\ref{fig:collisionComparison_LHD})-(\ref{fig:collisionComparison_W7X}).
